---
title: "QMS 101 Statistics"
subtitle: "Topic III: Simple  Regression and Correlation Analysis"
author: "Dr. Anna Fome"
format: 
  html:
    toc: true
    number-sections: true
    theme: cosmo
    math: mathjax
  pdf:
    documentclass: article
    toc: true
---

------------------------------------------------------------------------

# Introduction

In business, accounting, and finance, we often want to understand relationships between variables.

Examples:

-   Does advertising affect sales?
-   Does income affect spending?
-   Does price affect demand?

This topic helps us:

-   Understand relationships → **Correlation**
-   Predict values → **Regression**

# Scatter Diagrams

## Definition

A scatter diagram is a graph that shows the relationship between two variables.

Each point represents a pair of values (X, Y).

## R Code: Scatter Diagrams

### 1. Positive Relationship (Business Example: Advertising vs Sales)

```{r}
x <- c(10, 20, 30, 40, 50)
y <- c(100, 150, 200, 250, 300)
plot(x, y, main="Positive Relationship (Advertising vs Sales)", xlab="Advertising", ylab="Sales")
```

👉 Meaning: As advertising increases, sales also increase.

------------------------------------------------------------------------

### 2. Negative Relationship (Business Example: Price vs Demand)

```{r}
x <- c(10, 20, 30, 40, 50)
y <- c(500, 400, 300, 200, 100)
plot(x, y, main="Negative Relationship (Price vs Demand)", xlab="Price", ylab="Demand")
```

👉 Meaning: As price increases, demand decreases.

------------------------------------------------------------------------

### 3. No Relationship (Business Example: Shoe Size vs Salary)

```{r}
x <- c(10, 20, 30, 40, 50)
y <- c(200, 450, 300, 500, 350)
plot(x, y, main="No Relationship (Shoe Size vs Salary)", xlab="Shoe Size", ylab="Salary")
```

👉 Meaning: No clear relationship between the variables.

------------------------------------------------------------------------

# Correlation Analysis

## Definition

Correlation measures the strength and direction of a relationship between two variables.

-   Positive correlation → both variables move in the same direction
-   Negative correlation → variables move in opposite directions

------------------------------------------------------------------------

## Fully Solved Example (Business: Advertising and Sales)

A company records advertising and sales:

| Advertising (TZS ‘000) | Sales (TZS ‘000) |
|------------------------|------------------|
| 10                     | 100              |
| 20                     | 150              |
| 30                     | 200              |
| 40                     | 250              |
| 50                     | 300              |

This shows that as advertising increases, sales also increase.

👉 Without heavy calculation, we can observe a **strong positive relationship**.

Interpretation:

-   More advertising leads to higher sales

------------------------------------------------------------------------

# Simple Linear Regression

## Definition

Regression is used to **predict one variable using another**.

------------------------------------------------------------------------

## Model

$$Y = a + bX$$

Where:

-   Y = dependent variable (e.g. sales)
-   X = independent variable (e.g. advertising)
-   a = intercept
-   b = slope

------------------------------------------------------------------------

## Fully Solved Example (Business)

Using the same data:

Advertising (X): 10, 20, 30, 40, 50 Sales (Y): 100, 150, 200, 250, 300

### Step 1: Observe the pattern

When X increases by 10, Y increases by 50.

### Step 2: Find slope (b)

$$b = \frac{change\ in\ Y}{change\ in\ X} = \frac{50}{10} = 5$$

### Step 3: Find intercept (a)

Using a point (X=10, Y=100):

$$100 = a + 5(10)$$

$$100 = a + 50$$

$$a = 50$$

------------------------------------------------------------------------

### Regression Equation

$$Y = 50 + 5X$$

------------------------------------------------------------------------

## Forecasting Example

If advertising = 60:

$$Y = 50 + 5(60) = 350$$

👉 Predicted sales = 350

Interpretation:

-   Increasing advertising increases sales

------------------------------------------------------------------------

# Relationship Between Correlation and Regression

-   Correlation tells us **how strong** the relationship is
-   Regression tells us **how to predict** values

Simple idea:

-   Correlation = description
-   Regression = prediction

------------------------------------------------------------------------

# Multiple Regression (Introduction)

In real business situations, one factor is not enough.

Example:

-   Sales depend on advertising, price, and income

Model:

$$Y = a + b_1X_1 + b_2X_2$$

------------------------------------------------------------------------

# SPSS / R Commands

## Correlation in R

```{r}
cor(x, y)
```

## Regression in R

```{r}
model <- lm(y ~ x)
summary(model)
```

------------------------------------------------------------------------

# Summary (Simple Points)

-   Scatter diagrams show relationships visually
-   Correlation measures strength and direction
-   Regression is used for prediction
-   Positive relationship: both variables increase together
-   Negative relationship: one increases, the other decreases
-   No relationship: no clear pattern

------------------------------------------------------------------------

# Practice Questions

1.  Explain correlation in simple words
2.  Give a business example of a positive relationship
3.  What does regression help us do?
4.  Give an example of a negative relationship in business

------------------------------------------------------------------------

# Key Business Insight

-   Businesses use correlation to understand patterns
-   Businesses use regression to forecast sales, costs, and profits
-   Good decisions depend on understanding relationships between variables
